---
title: 'COVID-19 Confirmed Cases by Colorado County'
author: 'Joseph Tuccillo, University of Colorado-Boulder Department of Geography'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.path='figs/',
                      fig.pos='h',
                      fig.ext = 'pdf',
                      echo=FALSE,
                      warning=FALSE,
                      message=FALSE)
knitr::opts_knit$set(root.dir = '../')
```

## Daily Cases by Colorado County

```{r}
# helper functions for computing LQ and rates
source('code/support_functions.R')
```

```{r}
# load previous daily cases by county data
# fn <- paste0(paste('daily_cases_by_county', (Sys.Date() - 1), sep = '_'), '.RData')
fn <- max(list.files('data/legacy'))
fn <- file.path('data', 'legacy', fn)
load(fn)
```

```{r}
## append new cases ONLY if most recent previous date
## does not match the one-day lagged date
## OR it is past 4pm, when the new data is released
library(lubridate)

if(((Sys.Date() - 1) - max(daily_cases_cty$date) > 1) | (hour(Sys.time()) >= 16)){
    
    library(esri2sf)
    
    url <- 'https://services3.arcgis.com/66aUo8zsujfVXRIT/arcgis/rest/services/Colorado_COVID19_Positive_Cases/FeatureServer/0'
    
    cases_new <- esri2sf(url = url,
            outFields = c('LABEL', 'Date_Data_Last_Updated', 'County_Pos_Cases', 'County_Population'))
    
    ## strip geometry
    st_geometry(cases_new) <- NULL
    
    ## rename columns
    names(cases_new) <- c('attribute', 'date', 'value', 'pop')
    
    ## format date
    cases_new$date <- gsub('Data through ', '', cases_new$date)
    cases_new$date <- mdy(cases_new$date)
    
    # ensure county names are in alphabetical order (like legacy data)
    cases_new <- cases_new[match(sort(unique(daily_cases_cty$attribute)), cases_new$attribute),]
    
    # compute daily cases per 100k
    cases_new$cases_100k <- rate(cases_new)
    
    ## append `cases_new` to `daily_cases_cty`
    daily_cases_cty <- rbind(daily_cases_cty, cases_new)
    
}
```

```{r}
## remove NA cases (from "unknown" or "pending" counties)
daily_cases_cty <- daily_cases_cty[complete.cases(daily_cases_cty),]
```

```{r eval = F}
# sanity check - should display 64 counties per day when run (starting at 3/19)
lapply(split(daily_cases_cty, daily_cases_cty$date), nrow)
```

```{r eval = F}
## compute cases per 100k
daily_cases_cty$cases_100k <- rate(daily_cases_cty)
```

```{r}
## limit to nonzero cases/100k (for visualizations only)
daily_cases_cty_viz <- daily_cases_cty[daily_cases_cty$cases_100k > 0,]
```


### Raw Cases per 100,000 People by County

```{r}
current_date <- max(daily_cases_cty$date)
```


```{r daily-cases-100k, fig.height = 12, fig.width = 12, dpi = 320}
library(ggplot2)
trjplt1 <- ggplot(data = daily_cases_cty_viz, aes(x = date, y = cases_100k, color = attribute)) + 
    geom_line(size = 0.75) +
    facet_wrap(~attribute) +
    xlab('Date') +
    ylab('Cases per 100,000 People') +
    ggtitle(label = 'Colorado COVID-19 Daily Confirmed Case Trajectories',
                   subtitle = paste0('(', '2020-03-19 - ', current_date, ')')) +
    theme_bw() +
    theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
    theme(plot.title = element_text(hjust = 0.5), 
          plot.subtitle = element_text(hjust = 0.5), 
          legend.position = 'none')
trjplt1
```

### Log-Transformed Cases per 100,000 People by County

```{r daily-cases-100k-log, fig.height = 12, fig.width = 12}
# with logarithmic y-axis
trjplt2 <- ggplot(data = daily_cases_cty_viz, aes(x = date, y = cases_100k, color = attribute)) + 
    geom_line(size = 0.75) +
    facet_wrap(~attribute) +
    scale_y_log10() +
    xlab('Date') +
    ylab('Log-Transformed Cases per 100,000 People') +
    ggtitle(label = 'Colorado COVID-19 Daily Confirmed Case Trajectories',
                   subtitle = paste0('(', '2020-03-19 - ', current_date, ')')) +
    theme_bw() +
    theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
    theme(plot.title = element_text(hjust = 0.5), 
          plot.subtitle = element_text(hjust = 0.5), 
          legend.position = 'none')
trjplt2
```

```{r}
## write output images
ggsave(plot = trjplt1, filename = 'daily-cases-raw.png', path = 'figs/', device = 'png', dpi = 320, width = 12, height = 12, units = 'in')
ggsave(plot = trjplt2, filename = 'daily-cases-log.png', path = 'figs/', device = 'png', dpi = 320, width = 12, height = 12, units = 'in')
```


\newpage
## County Confirmed Case Trajectories (Experimental)


### Methodology

Use Affinity Propagation clustering to group daily reports of confirmed cases by county based on two criteria:

1. The current rate of confirmed cases per 100,000 people.
2. The change in confirmed cases per 100,000 people since the previous day.

- March 19, 2020 is used as the intial date, since it is the first day at which the change in cases/100k people by county can be measured (3/18/2020 marks the first day in which all counties were reporting). (_**NOTE** that the data is published with a one-day lag._)

- From the ensemble of daily clusterings, measure the percentage of days to date that any two counties shared a cluster label.
    - **EXPERIMENTAL**: the similarities are now upweighted by more recent dates. This is subject to change.

- Perform a final clustering (also using Affinity Propagation) to group the change trajectories from 3/19/2020 to present.

```{r}
# convert daily cases back to list format
daily_cases_split <- split(daily_cases_cty, daily_cases_cty$date)
```

```{r}
current_cases <- daily_cases_split[[length(daily_cases_split)]]
excl <- current_cases$attribute[current_cases$cases_100k == 0]
```


```{r}
# build cluster ensemble
source('code/apcluster_helper.R')
ens <- sapply(4:length(daily_cases_split), function(i){
    
    day1 <- daily_cases_split[[i-1]]
    day2 <- daily_cases_split[[i]]
    
    # lagged rate of change
    rate_of_change <- (day2$cases_100k - day1$cases_100k)
    # rate_of_change <- rate_of_change / day1$cases_100k # change proportional to day1 cases
    rate_of_change[is.na(rate_of_change)] <- 0 # occurs when d1: 0 cases/100k to d2: 0 cases/100k
    
    inx <- scale(cbind(rate_of_change, day2$cases_100k))
    rownames(inx) <- day2$attribute
    
    # exclude current zero cases
    inx <- inx[!rownames(inx) %in% excl,]
    
    set.seed(909) # so the labels stay consistent
    apcx <- apcluster(negDistMat(inx))
    
    clustx <- clust_vec(apcx)
    names(clustx) <- rownames(inx)
    clustx
    
})
```

```{r eval = F}
## simple weights
w <- 1:length(daily_cases_split) / length(daily_cases_split)
```

```{r}
# convert labels to binary format
# ed <- lapply(1:ncol(ens), function(i){
es <- lapply(1:ncol(ens), function(i){
    
    indmat <- model.matrix(~factor(ens[,i])-1)
    colnames(indmat) <- paste0(paste0('d', i), letters[1:ncol(indmat)])
    rownames(indmat) <- rownames(ens)
    d <- as.matrix(vegan::vegdist(indmat, method = 'jaccard'))
    s <- ifelse(d == 1, 0, 1)
    # s * w[i]
    # s * (1 + w[i])
    s
    
})
es <- Reduce(es, f = '+')
```

```{r}
set.seed(909) # so the labels stay consistent
apce <- apcluster(es)
```

```{r}
## cluster labels
clust <- clust_vec(apce)
clust <- sprintf('%02d', clust)
names(clust) <- rownames(es)
excl.pad <- rep('No Confirmed Cases', length(excl))
names(excl.pad) <- excl
clust <- c(clust, excl.pad)
```

```{r}
# load county boundaries
library(sf)
counties <- read_sf('data', 'cb_2018_us_county_500k')
counties <- counties[counties$STATEFP == '08',]
counties <- counties[match(names(clust), counties$NAME),]
```

```{r}
# assign cluster labels to counties
counties['cluster'] <- factor(clust)
```

```{r}
# append cluster labels
daily_cases_cty$cluster <- counties$cluster[match(daily_cases_cty$attribute, counties$NAME)]
daily_cases_cty_viz$cluster <- counties$cluster[match(daily_cases_cty_viz$attribute, counties$NAME)]
```

```{r}
## labels for overview map
county_labels <- st_coordinates(st_centroid(counties))
county_labels <- data.frame(GEOID = counties$NAME, county_labels)
```

```{r}
## color pallette
library(RColorBrewer)
pal <- c(brewer.pal(length(unique(clust[clust != 'No Confirmed Cases'])), 'Set3'), 'snow')
```


```{r}
## map plot
library(shadowtext)
mapplot <- ggplot() + 
            geom_sf(data = counties, aes(fill = cluster), size = 0.25) +
            # scale_fill_brewer(type = 'qual', palette = 'Set3') +
            scale_fill_manual(values = pal) +
            geom_shadowtext(data = county_labels, aes(x = X, y = Y, label = GEOID), color = 'white', size = 3, fontface = 'bold', stroke = 0.1) +
            ggtitle(label = 'Colorado COVID-19 Daily Confirmed Case Trajectories',
                   subtitle = paste0('(', '2020-03-19 - ', (Sys.Date() - 1), ')')) +
            theme_void() +
            theme(plot.title = element_text(hjust = 0.5), 
                  plot.subtitle = element_text(hjust = 0.5), 
                  legend.position = 'none')
```

```{r}
## change trajectories by cluster
# untransformed y-axis
trjplot <- ggplot(data = daily_cases_cty_viz[daily_cases_cty_viz$date >= '2020-03-19',],
            aes(x = date, y = cases_100k, color = cluster, group = attribute)) + 
          geom_line(size = 0.75) +
          # scale_color_brewer(type = 'qual', palette = 'Set3') +
          scale_color_manual(values = pal) +
          geom_smooth(data = daily_cases_cty_viz[daily_cases_cty_viz$date >= '2020-03-19',],
                      aes(x = date, y = cases_100k), inherit.aes = F, color = 'black', fill = NA,   
                      linetype = 'dashed', size = 0.5) + 
          facet_wrap(~cluster) +
          xlab('Date') +
          ylab('Cases per 100,000 People') +
          geom_vline(xintercept = 4) +
          theme_bw() +
          theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
          theme(legend.position = 'none')
```

```{r}
# with logarithmic y-axis
trjplot_log <- ggplot(data = daily_cases_cty_viz[daily_cases_cty_viz$date >= '2020-03-19',],
               aes(x = date, y = cases_100k, color = cluster, group = attribute)) + 
          geom_line(size = 0.75) +
          # scale_color_brewer(type = 'qual', palette = 'Set3') +
          scale_color_manual(values = pal) +
          geom_smooth(data = daily_cases_cty_viz[daily_cases_cty_viz$date >= '2020-03-19',],
                      aes(x = date, y = cases_100k), inherit.aes = F, color = 'black', fill = NA,  
                      linetype = 'dashed', size = 0.5) + 
          facet_wrap(~cluster) +
          scale_y_log10() +
          xlab('Date') +
          ylab('Log-Transformed Cases per 100,000 People') +
          geom_vline(xintercept = 4) +
          theme_bw() +
          theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
          theme(legend.position = 'none')
```

### Raw Cases per 100,000 people by cluster

```{r county-clusters-raw, fig.height = 13, fig.width = 8}
clsplt1 <- cowplot::plot_grid(mapplot, trjplot, ncol = 1)
clsplt1
ggsave(filename = 'county-clusters-raw.png', path = 'figs/', device = 'png', dpi = 320, width = 8, height = 13, units = 'in')
```

### Log-transformed Cases per 100,000 people by cluster

```{r county-clusters--log, fig.height = 13, fig.width = 8}
clsplt2 <- cowplot::plot_grid(mapplot, trjplot_log, ncol = 1)
clsplt2
ggsave(filename = 'county-clusters-log.png', path = 'figs/', device = 'png', dpi = 320, width = 8, height = 13, units = 'in')
```

## Save New Data

```{r}
# fnx <- paste0(paste('daily_cases_by_county', Sys.Date(), sep = '_'), '.RData')
fnx <- paste0(paste('daily_cases_by_county', (max(daily_cases_cty$date) + 1), sep = '_'), '.RData')
fnx <- file.path('data', 'legacy', fnx)
daily_cases_cty <- daily_cases_cty[,-ncol(daily_cases_cty)]  # drop `cluster` column 
save(daily_cases_cty, file = fnx)
```

