---
title: 'COVID-19 Confirmed Cases by Colorado County'
author: 'Joseph Tuccillo, University of Colorado-Boulder Department of Geography'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.path='figs/',
                      fig.pos='h',
                      # out.extra='',
                      fig.ext = 'pdf',
                      # fig.ext = 'png',
                      # dpi = 320,
                      echo=FALSE,
                      warning=FALSE,
                      message=FALSE)
knitr::opts_knit$set(root.dir = '../')
```

## Daily Cases by Colorado County

```{r}
# helper functions for computing LQ and rates
source('code/support_functions.R')
```

```{r eval = F}
## fetch any new case files
# will skip download if no new files are available
## THIS IS SKIPPED -- all CSV files are now treated as legacy data!
source('code/get_new_case_data.R')
```

```{r}
# county population (sex by age, aggregates to total county pop)
co_sxa <- read.csv('data/co_sex_by_age_by_county_acs_2018_5yr.csv',
                   stringsAsFactors = F) 

# subset the county ID and total population columns
co_pop <- co_sxa[,c('Geo_NAME', 'ACS18_5yr_B01001001')]

# rename total population column
names(co_pop)[2] <- 'pop'
```


```{r}
# some of the earlier dates don't have the ` County` suffix,
# so remove this from the total population count IDs
co_pop$Geo_NAME <- gsub(' County', '', co_pop$Geo_NAME)
```

```{r}
# all filenames in the `data` folder
# legacy files (CSV from google drive)
data_files <- list.files('data')
```

```{r}
# filenames for all (legacy) covid19 case data files
covid_sf <- data_files[startsWith(data_files, 'covid19')]
```

```{r}
## build daily case data
daily_cases_cty <- lapply(covid_sf, function(covx){
    
    # full path to the file
    path.covx <- file.path('data', covx)
    
    # extract date info from filename
    date.covx <- gsub('covid19_case_summary_', '', covx) # strip prefix
    date.covx <- gsub('.csv', '', date.covx) # strip file extension
    
    # read in file
    data.covx <- read.csv(path.covx, stringsAsFactors = F)
    
    # subset to counties
    cty.covx <- data.covx[data.covx$description == 'Case Counts by County' & data.covx$metric == 'Cases',] 
    
    # append date 
    cty.covx <- data.frame(date = date.covx, cty.covx, stringsAsFactors = F)
    
    # convert date from text to date/time
    cty.covx$date <- lubridate::date(cty.covx$date)
    
    # ensure `value` column is numeric
    cty.covx$value <- as.numeric(cty.covx$value)
    
    # strip `County` prefix if it exists 
    cty.covx$attribute <- gsub(' County', '', cty.covx$attribute)
    
    # append total population counts
    cty.covx <- merge(cty.covx, co_pop, by.x = 'attribute', by.y = 'Geo_NAME')
    
    # append location quotient and case rates
    # cty.covx$lq <- lq(cty.covx)
    cty.covx$cases_100k <- rate(cty.covx)
    
    # ensure county names in alphabetical order
    cty.covx <- cty.covx[match(sort(cty.covx$attribute), cty.covx$attribute),]
    
    # cty.covx[,c('attribute', 'date', 'value', 'pop', 'lq', 'cases_100k')]
    cty.covx[,c('attribute', 'date', 'value', 'pop', 'cases_100k')]
    
})
```

```{r}
# combine results into a single data frame
daily_cases_cty <- do.call(rbind, daily_cases_cty)
```

```{r}
## Read in the newly published case data file (geojson)
cases_new <- rgdal::readOGR('https://opendata.arcgis.com/datasets/fbae539746324ca69ff34f086286845b_0.geojson')
```

```{r}
## append new case data
cases_new <- with(cases_new@data, 
                  data.frame(
                      attribute = LABEL,
                      date = Sys.Date() - 1, 
                      value = County_Pos_Cases,
                      pop = County_Population,
                      stringsAsFactors = F
                  ))
```

```{r}
## compute cases/100k
cases_new$cases_100k <- rate(cases_new)
```

```{r}
# ensure county names are in alphabetical order (like legacy data)
cases_new <- cases_new[match(unique(daily_cases_cty$attribute), cases_new$attribute),]
```

```{r}
daily_cases_cty <- rbind(daily_cases_cty, cases_new)
```

```{r}
## limit to nonzero cases/100k
daily_cases_cty_viz <- daily_cases_cty[daily_cases_cty$cases_100k > 0,]
```


### Raw Cases per 100,000 People by County

```{r daily-cases-100k, fig.height = 12, fig.width = 12, dpi = 320}
library(ggplot2)
trjplt1 <- ggplot(data = daily_cases_cty_viz, aes(x = date, y = cases_100k, color = attribute)) + 
    geom_line(size = 0.75) +
    facet_wrap(~attribute) +
    xlab('Date') +
    ylab('Cases per 100,000 People') +
    theme_bw() +
    theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
    theme(legend.position = 'none')
trjplt1
```

### Log-Transformed Cases per 100,000 People by County

```{r daily-cases-100k-log, fig.height = 12, fig.width = 12}
# with logarithmic y-axis
trjplt2 <- ggplot(data = daily_cases_cty_viz, aes(x = date, y = cases_100k, color = attribute)) + 
    geom_line(size = 0.75) +
    facet_wrap(~attribute) +
    scale_y_log10() +
    xlab('Date') +
    ylab('Log-Transformed Cases per 100,000 People') +
    theme_bw() +
    theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
    theme(legend.position = 'none')
trjplt2
```

```{r}
## write output images
ggsave(plot = trjplt1, filename = 'daily-cases-raw.png', path = 'figs/', device = 'png', dpi = 320, width = 12, height = 12, units = 'in')
ggsave(plot = trjplt2, filename = 'daily-cases-log.png', path = 'figs/', device = 'png', dpi = 320, width = 12, height = 12, units = 'in')
```


\newpage
## County Confirmed Case Trajectories (Experimental)


### Methodology

Use Affinity Propagation clustering to group daily reports of confirmed cases by county based on two criteria:

1. The current rate of confirmed cases per 100,000 people.
2. The change in confirmed cases per 100,000 people since the previous day.

- March 20, 2020 is used as the intial date, since it is the first day at which the change in cases/100k people by county can be measured (3/19/2020 marks the first day in which all counties were reporting).

- From the ensemble of daily clusterings, measure the percentage of days to date that any two counties shared a cluster label.

- Perform a final clustering (also using Affinity Propagation) to group the change trajectories from 3/20/2020 to present.

```{r}
# convert daily cases back to list format
daily_cases_split <- split(daily_cases_cty, daily_cases_cty$date)
```

```{r}
current_cases <- daily_cases_split[[length(daily_cases_split)]]
excl <- current_cases$attribute[current_cases$cases_100k == 0]
```


```{r}
# build cluster ensemble
source('code/apcluster_helper.R')
ens <- sapply(4:length(daily_cases_split), function(i){
    
    day1 <- daily_cases_split[[i-1]]
    day2 <- daily_cases_split[[i]]
    
    # lagged rate of change
    rate_of_change <- (day2$cases_100k - day1$cases_100k)
    # rate_of_change <- rate_of_change / day1$cases_100k # change proportional to day1 cases
    rate_of_change[is.na(rate_of_change)] <- 0 # occurs when d1: 0 cases/100k to d2: 0 cases/100k
    
    inx <- scale(cbind(rate_of_change, day2$cases_100k))
    rownames(inx) <- day2$attribute
    
    # exclude current zero cases
    inx <- inx[!rownames(inx) %in% excl,]
    
    set.seed(909) # so the labels stay consistent
    apcx <- apcluster(negDistMat(inx))
    
    clustx <- clust_vec(apcx)
    names(clustx) <- rownames(inx)
    clustx
    
})
```

```{r}
# convert labels to binary format
ed <- lapply(1:ncol(ens), function(i){
    
    indmat <- model.matrix(~factor(ens[,i])-1)
    colnames(indmat) <- paste0(paste0('d', i), letters[1:ncol(indmat)])
    rownames(indmat) <- rownames(ens)
    as.matrix(vegan::vegdist(indmat, method = 'jaccard'))
    
})
ed <- Reduce(ed, f = '+')
ed <- ed / length(daily_cases_split)
```

```{r}
## build similarity matrix 
# similarity as percentage of times two counties co-clustered
es <- 1 - ed
```

```{r}
set.seed(909) # so the labels stay consistent
apce <- apcluster(es)
```

```{r}
## cluster labels
clust <- clust_vec(apce)
clust <- sprintf('%02d', clust)
names(clust) <- rownames(es)
excl.pad <- rep('No Confirmed Cases', length(excl))
names(excl.pad) <- excl
clust <- c(clust, excl.pad)
```

```{r}
# load county boundaries
library(sf)
counties <- read_sf('data', 'cb_2018_us_county_500k')
counties <- counties[counties$STATEFP == '08',]
counties <- counties[match(names(clust), counties$NAME),]
```

```{r}
# assign cluster labels to counties
counties['cluster'] <- factor(clust)
```

```{r}
daily_cases_cty$cluster <- counties$cluster[match(daily_cases_cty$attribute, counties$NAME)]
daily_cases_cty_viz$cluster <- counties$cluster[match(daily_cases_cty_viz$attribute, counties$NAME)]
```

```{r}
## labels for overview map
county_labels <- st_coordinates(st_centroid(counties))
county_labels <- data.frame(GEOID = counties$NAME, county_labels)
```

```{r}
## color pallette
library(RColorBrewer)
pal <- c(brewer.pal(length(unique(clust[clust != 'No Confirmed Cases'])), 'Set3'), 'snow')
```


```{r}
## map plot
library(shadowtext)
mapplot <- ggplot() + 
            geom_sf(data = counties, aes(fill = cluster), size = 0.25) +
            # scale_fill_brewer(type = 'qual', palette = 'Set3') +
            scale_fill_manual(values = pal) +
            geom_shadowtext(data = county_labels, aes(x = X, y = Y, label = GEOID), color = 'white', size = 3, fontface = 'bold', stroke = 0.1) +
            ggtitle(label = 'Colorado COVID-19 Daily Confirmed Case Trajectories',
                   subtitle = paste0('(', '2020-03-20 - ', (Sys.Date() - 1), ')')) +
            theme_void() +
            theme(plot.title = element_text(hjust = 0.5), 
                  plot.subtitle = element_text(hjust = 0.5), 
                  legend.position = 'none')
```

```{r}
## change trajectories by cluster
# untransformed y-axis
trjplot <- ggplot(data = daily_cases_cty_viz[daily_cases_cty_viz$date >= '2020-03-20',],
            aes(x = date, y = cases_100k, color = cluster, group = attribute)) + 
          geom_line(size = 0.75) +
          # scale_color_brewer(type = 'qual', palette = 'Set3') +
          scale_color_manual(values = pal) +
          geom_smooth(data = daily_cases_cty_viz[daily_cases_cty_viz$date >= '2020-03-20',],
                      aes(x = date, y = cases_100k), inherit.aes = F, color = 'black', fill = NA,   
                      linetype = 'dashed', size = 0.5) + 
          facet_wrap(~cluster) +
          xlab('Date') +
          ylab('Cases per 100,000 People') +
          geom_vline(xintercept = 4) +
          theme_bw() +
          theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
          theme(legend.position = 'none')
```

```{r}
# with logarithmic y-axis
trjplot_log <- ggplot(data = daily_cases_cty_viz[daily_cases_cty_viz$date >= '2020-03-20',],
               aes(x = date, y = cases_100k, color = cluster, group = attribute)) + 
          geom_line(size = 0.75) +
          # scale_color_brewer(type = 'qual', palette = 'Set3') +
          scale_color_manual(values = pal) +
          geom_smooth(data = daily_cases_cty_viz[daily_cases_cty_viz$date >= '2020-03-20',],
                      aes(x = date, y = cases_100k), inherit.aes = F, color = 'black', fill = NA,  
                      linetype = 'dashed', size = 0.5) + 
          facet_wrap(~cluster) +
          scale_y_log10() +
          xlab('Date') +
          ylab('Log-Transformed Cases per 100,000 People') +
          geom_vline(xintercept = 4) +
          theme_bw() +
          theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
          theme(legend.position = 'none')
```

### Raw Cases per 100,000 people by cluster

```{r county-clusters-raw, fig.height = 13, fig.width = 8}
clsplt1 <- cowplot::plot_grid(mapplot, trjplot, ncol = 1)
clsplt1
ggsave(filename = 'county-clusters-raw.png', path = 'figs/', device = 'png', dpi = 320, width = 8, height = 13, units = 'in')
```

### Log-transformed Cases per 100,000 people by cluster

```{r county-clusters--log, fig.height = 13, fig.width = 8}
clsplt2 <- cowplot::plot_grid(mapplot, trjplot_log, ncol = 1)
clsplt2
ggsave(filename = 'county-clusters-log.png', path = 'figs/', device = 'png', dpi = 320, width = 8, height = 13, units = 'in')
```
